Searching for Software in NSF Awards

### 2023-02-10 paper/paper.qmd
More paper progress and EDA section start

## Data Collection

### Finding Software Produced by NSF Awards

@@

Markdown(
  f"Our final round of trial coding showed that there was almost perfect agreement "
  f"between the research team (κ={round(irr_kappa, 3)})."
)
```

@@

Our final step in data collection was to link the annotated GitHub repositories back to specific NSF awards. To do so, we created a script which would load the webpage for each GitHub repository, scrape the content of the README and find the specific NSF award ID number(s) referenced. While annotating the dataset, and with this script, our dataset size was reduced as we found some repositories were returned in the initial search because of the "NSF" acronym being used by other, non-United-States governmental agencies which also fund research.

@@

We trained three different models using the text from either the award abstract or project outcomes report. The models trained include a logistic regression model trained with TF-IDF word embeddings (`tfidf-logit`), a logistic regression model trained with semantic embeddings (`semantic-logit`), and a fine-tuned transformer (`fine-tuned-transformer`). The semantic embeddings and the base model from which we fine-tuned our own transformer model was the `distilbert-base-uncased-finetuned-sst-2-english` model [@hf_canonical_model_maintainers_2022]. Additionally, we evaluated a regular expression which matched on common terms relating to software as a baseline.

@@

model_results = fit_and_eval_all_models()
```

@@

```{python}

@@

@@

The dataset, including our software prediction values is available via a link on our GitHub README.

@@

## Discussion

Our initial method for attempting to find research software was to first use data from the NSF API then look through the abstracts, project outcomes reports, and any attached papers to find references to code repositories. However as this proved to be such a manual and time consuming process we decided against it. While attempting to use this NSF first method to create the dataset, we found that many awards and papers that reference computational methods still do not provide a reference web link to their code repositories or websites. In some cases we manually found repositories related to an award or paper by manual search ourselves. While we support including references to code repositories in award abstracts, outcomes reports, and papers, future research should be conducted on how to enable reconnection of papers with code.

As our dataset was entirely composed of NSF awards which ultimately released or hosted software on GitHub, it is possible that our models learned not to predict if an NSF award would produce software, but rather, if an NSF award would produce “software hosted on GitHub.”

@@

>>>>

## Data Collection and Annotation {#sec-data-collection}

### Finding Software Produced by NSF Awards {#sec-finding-soft}

@@

Markdown(
  f"Our final round of trial coding showed that there was almost perfect agreement "
  f"between the research team (K={round(irr_kappa, 3)})."
)
```

@@

Our final step in the data collection process was to link the annotated GitHub repositories back to specific NSF awards. To do so, we created a script which would load the webpage for each GitHub repository, scrape the content of the README and find the specific NSF award ID number(s) referenced. While annotating the dataset, and with this script, our dataset size was reduced as we found some repositories were returned in the initial search because of the "NSF" acronym being used by other, non-United-States governmental agencies which also fund research.

@@

We trained three different models using the text from either the award abstract or project outcomes report. The models trained include a logistic regression model trained with TF-IDF word embeddings (`tfidf-logit`), a logistic regression model trained with semantic embeddings (`semantic-logit`), and a fine-tuned transformer (`fine-tuned-transformer`). The semantic embeddings and the base model from which we fine-tuned our own transformer model was the `distilbert-base-uncased-finetuned-sst-2-english` model [@hf_canonical_model_maintainers_2022]. Additionally, we evaluated a regular expression which matched on common terms relating to software as a baseline. Each model was trained with 80% of the data. We then test each of the models and use F1 to rank each models performance.

@@

model_results = fit_and_eval_all_models(train_transformer=False)
```

@@

@@

While the models trained with the project outcomes reports were trained with less data, the best model of the group achieved a higher F1 than any of the models trained with the abstracts. While we have not investigated further, we believe this to be because the project outcomes reports contain more direct citation of produced software rather than an abstract's promise of software production.

The data used for training, and functions to reproduce these models, are made available via our Python package: [`soft-search`](https://github.com/si2-urssi/eager).

@@

We have compiled the NSF Soft(ware) Search Dataset (`NSF-Soft-Search`). The NSF-Soft-Search dataset contains the metadata, abstract text, project outcomes report text, for all NSF awarded projects during the 2010-2023 period. The NSF-Soft-Search dataset additionally contains our predictions for software production using both texts respectively.

```{python}
#| code-fold: true
#| warning: false
#| label: tbl-nsf-soft-search-stats
#| tbl-cap: Composition of the NSF Soft Search Dataset

import pandas as pd

full_dataset = pd.read_parquet("nsf-soft-search-2022.parquet")

# Get descriptive statistics for each major program
desc_table_rows = []
for program in full_dataset.majorProgram.unique():
    subset = full_dataset[full_dataset.majorProgram == program]
    total_awards_funded = len(subset)
    will_produce_software = subset[
      subset.prediction_from_abstract == "software-predicted"
    ]
    num_awards_will_produce_software = len(will_produce_software)
    percent_awards_will_produce = num_awards_will_produce_software / total_awards_funded

    desc_table_rows.append({
        "Funding Program": program,
        "# Awards Funded": total_awards_funded,
        "# Awards 'Produce Software'": num_awards_will_produce_software,
        "% Awards 'Produce Software'": percent_awards_will_produce,
    })

desc_table = pd.DataFrame(desc_table_rows).sort_values(
    by=["# Awards Funded", "% Awards 'Produce Software'"],
    ascending=False,
).reset_index(drop=True)
desc_table
```

@@

* Trends in software production over time from abstract
```{python}
#| code-fold: true
#| warning: false

# Create start year, experation year, and award duration columns for downstream
full_dataset["startYear"] = full_dataset.startDate.apply(lambda d: d.split("/")[-1])
full_dataset["expYear"] = full_dataset.expDate.apply(lambda d: d.split("/")[-1])
full_dataset["awardDuration"] = full_dataset.apply(
  lambda r: int(r.expYear) - int(r.startYear),
  axis=1,
)

# Subset the dataset to remove programs which either do not have many awards in total
# or do not produce awards which result in software production
high_soft_prog_df = full_dataset[~full_dataset.majorProgram.isin(
  ["EHR", "SBE", "TIP", "OISE", "OIA"]
)]
```

```{python}
#| code-fold: true
#| warning: false
#| label: fig-soft-over-time
#| fig-cap: Software Production Over Time (Using Predictions from Abstracts)

import seaborn as sns
import matplotlib.pyplot as plt
sns.set_context("paper")
sns.set_theme(
    style="ticks", rc={"axes.spines.right": False, "axes.spines.top": False}
)
sns.set_style("darkgrid", {"grid.color": "#000000", "grid.linestyle": ":"})
sns.set_palette(
    [
        "#ff6a75",  # cherry-red
        "#0060df",  # ocean-blue
        "#068989",  # moss-green
        "#712290",  # purple
        "#FFA537",  # orange
        "#FF2A8A",  # pink
        "#9059FF",  # lavender
        "#00B3F5",  # light-blue / sky-blue
        "#005e5e",  # dark blueish green
        "#C50143",  # dark-red / maroon
        "#3fe1b0",  # seafoam / mint
        "#063F96",  # dark-blue / navy-blue
        "#FFD567",  # banana-yellow
    ]
)

# Prep data for plotting software production over time
soft_produced_over_time_rows = []
for start_year in high_soft_prog_df.startYear.unique():
    year_data = high_soft_prog_df[high_soft_prog_df.startYear == start_year]
    for program_name, award_count in year_data.majorProgram.value_counts().items():
        program_preds = year_data[
          year_data.majorProgram == program_name
        ].prediction_from_abstract.value_counts()
        try:
            predicted_count = program_preds.loc["software-predicted"]
        except KeyError:
            predicted_count = 0

        soft_produced_over_time_rows.append({
            "Funding Program": program_name,
            "Award Start Year": start_year,
            "% Awards 'Produce Software'": predicted_count / award_count,
        })

soft_produced_over_time = pd.DataFrame(soft_produced_over_time_rows)

# Plot over time
sns.barplot(
    data=soft_produced_over_time,
    x="Award Start Year",
    y="% Awards 'Produce Software'",
    hue="Funding Program",
    order=[
        str(i)
        for i in range(
            high_soft_prog_df.startYear.astype(int).min(),
            high_soft_prog_df.startYear.astype(int).max(),
        )
    ],
)
plt.xticks(rotation=45, ha="right")
plt.tight_layout()
```

```{python}
#| code-fold: true
#| warning: false
#| label: fig-soft-over-duration
#| fig-cap: Software Production Grouped By Award Duration (Using Predictions from Abstracts)

# Prep data for plotting software production over time
soft_produced_over_duration_rows = []
for award_duration in high_soft_prog_df.awardDuration.unique():
    duration_data = high_soft_prog_df[high_soft_prog_df.awardDuration == award_duration]
    for program_name, award_count in duration_data.majorProgram.value_counts().items():
        program_preds = duration_data[
          duration_data.majorProgram == program_name
        ].prediction_from_abstract.value_counts()
        try:
            predicted_count = program_preds.loc["software-predicted"]
        except KeyError:
            predicted_count = 0

        soft_produced_over_duration_rows.append({
            "Funding Program": program_name,
            "Award Duration (Years)": award_duration,
            "% Awards 'Produce Software'": predicted_count / award_count,
        })

soft_produced_over_duration = pd.DataFrame(soft_produced_over_duration_rows)

# Plot over time
sns.barplot(
    data=soft_produced_over_duration,
    x="Award Duration (Years)",
    y="% Awards 'Produce Software'",
    hue="Funding Program",
)
plt.tight_layout()
```

* Number of examples that initially were predicted as producing software but didn't over time
* Number of examples that initially were not predicted to produce software but did over time

## Conclusion

We introduce NSF-Soft-Search, a pair of novel datasets for studying software production from NSF funded projects. The NSF-Soft-Search GitHub Annotations dataset is used to train models which aim to predict software production from either the NSF award abstract text or the project outcomes report text. We used these models to generate the NSF-Soft-Search Inferred dataset. The NSF-Soft-Search Inferred dataset includes project metadata, abstract, project outcomes report, and a predictions of software production for each NSF funded project between 2010 and 2023. We hope that NSF-Soft-Search helps further new studies and findings in understanding the role software development plays in scholarly publication.

All datasets and predictive models produced by this work are available from our GitHub repository: [`si2-urssi/eager`](https://github.com/si2-urssi/eager).

### Limitations

As discussed in @sec-data-collection, the NSF-Soft-Search GitHub Annotations dataset was entirely composed of NSF awards which ultimately released or hosted software (and other research products) on GitHub. Due to data collection strategy, it is possible that each of the predictive models learned not to predict if an NSF award would produce software, but rather, if an NSF award would produce “software hosted on GitHub.”

### Future Work

As discussed in @sec-finding-soft, our initial method for attempting to find research software produced from NSF supported awards was to search for references and promises of software production in the abstract, project outcomes report, and attached papers of each award. While attempting this approach to create the dataset, we found that many awards and papers that reference computational methods do not provide a reference web link to their code repositories or websites. In some cases, we found repositories related to an award or paper via Google and GitHub search ourselves. While we support including references to code repositories in award abstracts, outcomes reports, and papers, future research should be conducted on how to enable reconnection of papers with code.

@@

We wish to thank ....

This work was supported in part by NSF Grant
